---
title: "DATA 607 assignment 1:Loading Data into a Data Frame"
author: "Lionel jr. desroses"
output: html_document
date: "2026-09-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
```

## Introduction
for this project I plan on researching the correlation between higher income and graduated level. i Plan on finding a data set on students application for school over the pass years and also checking the amount number of graduate while comparing the different data set with wages_by_education. this would give a great explanation on the importance of higher education in different race. 

##What data challenges do you anticipate?
I anticipating many challenge through-out this transformation process. as you can see my dataset is a completely mess and it would take a lot of work to tune these data and make them tell a story for me. I do plans on fixing and tuning the data.




```{r wages_by_edu}
library(readr)
wage_by_edu <- read_csv("https://raw.githubusercontent.com/lioneljr17/LDATA607/refs/heads/main/wage_by_edu.csv")
names(wage_by_edu)

#library(readr)
#cc_state_sector_grads <- read_csv("https://raw.githubusercontent.com/lioneljr17/LDATA607/refs/heads/main/cc_state_sector_grads.csv")
#library(readr)
#graduates <- read_csv("https://raw.githubusercontent.com/lioneljr17/LDATA607/refs/heads/main/graduates.csv")
```




``` {r framing Data-wage_by_LevelOfEdu }
wage_by_LevelOfEdu <- wage_by_edu%>%
  select(year,less_than_hs,high_school,some_college,bachelors_degree,advanced_degree)%>%
  rename (Year =year, No_Diploma=less_than_hs,High_school_diploma=high_school, associates_degree =some_college, Bachelors_Degree =bachelors_degree, Graduate_School=advanced_degree)

head (wage_by_LevelOfEdu)
summary(wage_by_LevelOfEdu)

```

```{r framing Data-wage_by_LevelOfEdu_gender}
wage_by_LevelOfEdu_gender <- wage_by_edu%>%
  select (year,men_less_than_hs,men_high_school,men_some_college,men_bachelors_degree,men_advanced_degree,women_less_than_hs,women_high_school,women_some_college,women_bachelors_degree,women_advanced_degree)

head (wage_by_LevelOfEdu_gender)
summary(wage_by_LevelOfEdu_gender)

```

```{r framing Data -wage_by_LevelOfEdu_race}
wage_by_LevelOfEdu_race <- wage_by_edu%>%
  select (year,white_less_than_hs,white_high_school,white_some_college,white_bachelors_degree,white_advanced_degree,black_less_than_hs,black_high_school,black_some_college,black_bachelors_degree,black_advanced_degree,hispanic_less_than_hs,hispanic_high_school,hispanic_some_college,hispanic_bachelors_degree,hispanic_advanced_degree)

head (wage_by_LevelOfEdu_race)
summary(wage_by_LevelOfEdu_race)

```

```{r trying out graph}
#plot(wage_by_edu)
ggplot()+ geom_line(data = wage_by_edu, aes(x = year, y = less_than_hs,color ="Less High")) + 
  geom_line(data= wage_by_edu,aes(x = year, y = high_school, color="High School"))+
  geom_line(data= wage_by_edu,aes(x = year, y = some_college, color="Some College"))+
  geom_line(data= wage_by_edu, aes(x = year, y =bachelors_degree, color ="Bachelor's Degree")) +
  geom_line(data= wage_by_edu,aes(x = year, y =advanced_degree,color ="Advanced Degree"))+ 
  labs(title ="Average Wages By Education level",x ="Years",y="Average wages")+
   scale_color_manual(values = c(
     "Less High" = 'black',
      "High School" = 'red',
      "Some College" = 'blue',
      "Bachelor's Degree" = 'green',
      "Advanced Degree"='orange'
    )
  ) +theme_minimal()


```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
