---
title: "DATA 607 assignment 1:Loading Data into a Data Frame"
author: "Lionel jr. desroses"
output: html_document
date: "2026-09-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
```

## Introduction
for this project I plan on researching the correlation between higher income and graduated level. i Plan on finding a data set on students application for school over the pass years and also checking the amount number of graduate while comparing the different data set with wages_by_education. this would give a great explanation on the importance of higher education in different race. 

##What data challenges do you anticipate?
I anticipating many challenge through-out this transformation process. as you can see my dataset is a completely mess and it would take a lot of work to tune these data and make them tell a story for me. I do plans on fixing and tuning the data.




```{r wages_by_edu}
library(readr)
wage_by_edu <- read_csv("https://raw.githubusercontent.com/lioneljr17/LDATA607/refs/heads/main/wage_by_edu.csv")
library(readr)
cc_state_sector_grads <- read_csv("https://raw.githubusercontent.com/lioneljr17/LDATA607/refs/heads/main/cc_state_sector_grads.csv")
library(readr)
graduates <- read_csv("https://raw.githubusercontent.com/lioneljr17/LDATA607/refs/heads/main/graduates.csv")
#View(graduates)
#View(cc_state_sector_grads)
#View(wage_by_edu)
#summary (wage_by_edu) 
#summary (cc_state_sector_grads)
#summary (graduates)
names(wage_by_edu)

```

## Including Plots

You can also embed plots, for example:


```{r pressure, echo=FALSE}
#plot(wage_by_edu)
ggplot()+ geom_line(data = wage_by_edu, aes(x = year, y = some_college)) + 
  geom_line(aes(x = year, y = high_school), data= wage_by_edu, colour = 'red')+
  geom_line(aes(x = year, y =bachelors_degree), data= wage_by_edu, colour = 'green')


```


``` {r framing Data }
wage_by_LevelOfEdu <- wage_by_edu%>%
  select(year,less_than_hs,high_school,some_college,bachelors_degree,advanced_degree)%>%
  rename (Year =year, No_Diploma=less_than_hs,High_school_diploma=high_school, associates_degree =some_college, Bachelors_Degree =bachelors_degree, Graduate_School=advanced_degree)

head (wage_by_LevelOfEdu)
summary(wage_by_LevelOfEdu)
names(wage_by_edu)
```

```{r framing Data 2}
wage_by_LevelOfEdu_gender <- wage_by_edu%>%
  select (year,men_less_than_hs,men_high_school,men_some_college,men_bachelors_degree,men_advanced_degree,women_less_than_hs,women_high_school,women_some_college,women_bachelors_degree,women_advanced_degree)

```

```{r framing Data 2}
wage_by_LevelOfEdu_race <- wage_by_edu%>%
  select (year,white_less_than_hs,white_high_school,white_some_college,white_bachelors_degree,white_advanced_degree,black_less_than_hs,black_high_school,black_some_college,black_bachelors_degree,black_advanced_degree,hispanic_less_than_hs,hispanic_high_school,hispanic_some_college,hispanic_bachelors_degree,hispanic_advanced_degree)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
