---
title: "DATA 607 assignment 2B:Classification Metrics"
author: "Lionel jr. desroses"
output: html_document
editor: visual
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
date: "2026-09-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

## Introduction
For this project, I plan to focus on evaluating the performance of a binary classification model using a dataset of penguin observations. The goal is to use the model to predict whether a penguin belongs to the female class. Instead of relying only on the model's predicted class labels, I plan to analyze the predicted probabilities and examine how different probability thresholds affect the model's decisions.
My approach will begin by exploring the dataset to better understand the distribution of the target variable (sex) and determine whether the classes are balanced. I will also calculate the null error rate to establish a baseline for model performance. Next, I will create new class predictions using probability thresholds of 0.2, 0.5, and 0.8. For each threshold, I will build a confusion matrix to identify true positives, false positives, true negatives, and false negatives. Using these values, I will calculate accuracy, precision, recall, and F1 score. Comparing these results will help illustrate how threshold selection influences classification performance and the tradeoff between false positives and false negatives.

##load library and data 
In this section I am loading the data into R so i could be able to read it in my Rm and be able to modified it.
```{r import data set }
library(readr)
penguin_predictions <- read_csv("https://raw.githubusercontent.com/acatlin/data/refs/heads/master/penguin_predictions.csv")
View(penguin_predictions)
head((penguin_predictions))
```

## Exploratory Data Analysis
in this section, we are calling on the count function to count how many female and male we have in the sex column, we also created a visual graph to see the difference. by doing this we get to see if the classes are balances, as the data show they are not. based on the data we notice that we have more male than female in the dataset, and that male are the majority in the table 
```{r Data Analysis}
count (penguin_predictions,sex)

ggplot(penguin_predictions, aes(x = sex, fill = sex)) +
 geom_bar() +
 labs(
  title = "Distribution of Actual Classes",
  x = "Sex",
  y = "Count"
 ) +
 theme_minimal()
```

## Null Error Rate
In this section we are calculate the null error rate, the null error rate is usually represent the error produced by the model. this provide a baseline that our clasification model should outperform
```{r Null Error Rate}
table (penguin_predictions$sex)


1-(max(table (penguin_predictions$sex))/sum(table(penguin_predictions$sex)))
```

## Confusion Matrices
in this section we created a prediction colunm at each this threshhold 
```{r Confusion Matrices}

#threshhold (0.2,0.5,0.8)
penguin_predictions <- penguin_predictions %>% 
  mutate(pred_02 = ifelse(.pred_female>0.2, "female","male"),
         pred_05 = ifelse(.pred_female>0.5, "female","male"),
         pred_08 = ifelse(.pred_female>0.8, "female","male"))

head(penguin_predictions)


table (penguin_predictions$pred_02)
table (penguin_predictions$pred_05)
table (penguin_predictions$pred_08)


```

## Performance Metrics


# Treshold_0.2
in this section, we started assigning the, TP,FP,TN,FN for treshhold 0.2. At the 0.2 threshold, the model correctly identified 37 female penguins and 48 male penguins. The model produced 6 false positives and 2 false negatives.
```{r Treshold_0.2}
table(
  Actual= penguin_predictions$sex,
  predicted = penguin_predictions$pred_02
)
treshhold =0.2
treshhold_0.2 <- data.frame(
  metric =c("TP","FP","TN","FN"),
  value = c(37,6,48,2)
)
#treshhold_0.2

TP <- treshhold_0.2$value[treshhold_0.2$metric == "TP"]
FP <- treshhold_0.2$value[treshhold_0.2$metric == "FP"]
TN <- treshhold_0.2$value[treshhold_0.2$metric == "TN"]
FN <- treshhold_0.2$value[treshhold_0.2$metric == "FN"]

metrics_02 <- data.frame(
 Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
 Value = round(c(
  (TP + TN) / (TP + TN + FP + FN),
  TP / (TP + FP),
  TP / (TP + FN),
  2 * ((TP / (TP + FP)) * (TP / (TP + FN))) /
   ((TP / (TP + FP)) + (TP / (TP + FN)))
 ), 4)
)

metrics_02



```


# Treshold_0.5 
in this section, we started assigning the, TP,FP,TN,FN for treshhold 0.5,At the 0.5 threshold, the model correctly identified 36 female penguins and 51 male penguins. The number of false positives decreased to 3 while false negatives increased slightly to 3.
```{r Treshold_0.5}
table(
  Actual= penguin_predictions$sex,
  predicted = penguin_predictions$pred_05
)

treshhold =0.5
treshhold_0.5 <- data.frame(
  metric =c("TP","FP","TN","FN"),
  value = c(36,3,51,3)
)

TP <- treshhold_0.5$value[treshhold_0.5$metric == "TP"]
FP <- treshhold_0.5$value[treshhold_0.5$metric == "FP"]
TN <- treshhold_0.5$value[treshhold_0.5$metric == "TN"]
FN <- treshhold_0.5$value[treshhold_0.5$metric == "FN"]

metrics_05 <- data.frame(
 Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
 Value = round(c(
  (TP + TN) / (TP + TN + FP + FN),
  TP / (TP + FP),
  TP / (TP + FN),
  2 * ((TP / (TP + FP)) * (TP / (TP + FN))) /
   ((TP / (TP + FP)) + (TP / (TP + FN)))
 ), 4)
)

metrics_05

```

# Treshold_0.8 
in this section, we started assigning the, TP,FP,TN,FN for treshhold 0.8, At the 0.8 threshold, the model produced the fewest false positives (2) while maintaining the same number of false negatives as the 0.5 threshold.
```{r Treshold_0.8}

table(
  Actual= penguin_predictions$sex,
  predicted = penguin_predictions$pred_08
)

treshhold =0.8
treshhold_0.8 <- data.frame(
  metric =c("TP","FP","TN","FN"),
  value = c(36,2,52,3)
)

 
TP <- treshhold_0.8$value[treshhold_0.8$metric == "TP"]
FP <- treshhold_0.8$value[treshhold_0.8$metric == "FP"]
TN <- treshhold_0.8$value[treshhold_0.8$metric == "TN"]
FN <- treshhold_0.8$value[treshhold_0.8$metric == "FN"]

metrics_08 <- data.frame(
 Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
 Value = round(c(
  (TP + TN) / (TP + TN + FP + FN),
  TP / (TP + FP),
  TP / (TP + FN),
  2 * ((TP / (TP + FP)) * (TP / (TP + FN))) /
   ((TP / (TP + FP)) + (TP / (TP + FN)))
 ), 4)
)

metrics_08

```

## Threshold Analysis

The results demonstrate how changing the probability threshold affects classification performance. At the 0.2 threshold, the model achieved the highest recall (94.87%), meaning it identified most female penguins. However, this came at the cost of more false positives. As the threshold increased to 0.8, the number of false positives decreased, resulting in the highest precision (94.74%) and highest overall accuracy (94.62%). These results illustrate the tradeoff between identifying positive cases and minimizing incorrect positive predictions.
```{r}
metrics <- data.frame(
 Threshold = c("0.2", "0.5", "0.8"),
 Accuracy = c(0.9140, 0.9355, 0.9462),
 Precision = c(0.8605, 0.9231, 0.9474),
 Recall = c(0.9487, 0.9231, 0.9231),
 F1_Score = c(0.9024, 0.9231, 0.9351)
)

metrics
```


## Real-World Use Cases

Threshold 0.2
A low threshold may be preferred in medical screening applications where missing a positive case could have serious consequences. In this situation, maximizing recall is more important than minimizing false positives.
Threshold 0.8
A high threshold may be preferred in spam filtering or financial approval systems where false positives are costly. In these cases, higher precision is desirable even if some positive cases are missed.


## Conclusion

This project evaluated a binary classification model using different probability thresholds. The null error rate provided a baseline for comparison, while confusion matrices and performance metrics demonstrated how threshold selection affects model performance. The results showed that lower thresholds generally improve recall, while higher thresholds improve precision. The analysis highlights the importance of selecting an appropriate threshold based on the goals and requirements of a specific application.


