---
title: "607 peoject1"
author: "Lionel jr desroses"
output: html_document
date: "2026-09-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
for this project I plan on researching the correlation between higher income and graduated level. i Plan on finding a data set on students application for school over the pass years and also checking the amount number of graduate while comparing the different data set with wages_by_education. this would give a great explanation on the importance of higher education in different race. 

##What data challenges do you anticipate?
I anticipating many challenge through-out this transformation process. as you can see my dataset is a completely mess and it would take a lot of work to tune these data and make them tell a story for me. I do plans on fixing and tuning the data. 




```{r wages_by_edu}
summary(wage_by_edu)
summary (cc_state_sector_grads)
```

## Including Plots

You can also embed plots, for example:


```{r pressure, echo=FALSE}
plot(wage_by_edu)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
